Sequential rank agreement methods for comparison of ranked lists

### 2016-07-11 latex/superranker-manuscript.tex

The agreement of the lists regarding the rank given to an item $X_p$
can be measured by the variance across the lists
\begin{align}
  A(X_p) &= \E_Q\left[\left(R(X_P)-\E_Q R(X_P)\right)^2\right]\\
   &= \sum_{r=1}^{P!} \left(r(X_P)-\E_Q R(X_P)\right)^2 Q(r),\nn
\end{align}
\textsf{Ikke så vild med notationen her ovenfor}
where $\E_Q R(X_p)=\sum_{r=1}^{P!} r(X_p)Q(r)$ and $Q$ is a probability mass function
on the space of all permutations of lists with $P$ elements.
The empirical counterpart is
\begin{align}
\widehat{A}_L(X_p) = \frac{1}{L-1}\sum_{i=1}^L (R_i(X_p) - \overline{R}_L(X_p))^2, \quad \overline{R}_L(X_p) = \frac{1}{L}\sum_{i=1}^L R_i(X_p).\label{eq:empDistance}
\end{align}
For each item, the function $\widehat{A}_L$ has an interpretation as the expected
Euclidean distance of the individual rankings from the expected
ranking over the $L$ lists, and it corresponds to the same measure
that is used in method comparison studies to compute the limits of
agreement \citep{alt:bland:1983}.

For an integer $1\le d\le P$ we define the expected set of unique
items found by merging the first $d$ elements across the possible
lists:
\begin{align}
S(d) &= \left\{X_p; \left(\sum_{r=1}^{P!} 1\left(r(X_p)\le d\right) Q(r)\right) >0\right\}\label{eq:sumset}
\end{align}
where $1(\cdot)$ denotes the indicator function.
The empirical counterpart is the set of unique items ranked less than or
equal to $d$ in any of the $L$ lists:
\begin{equation}
\widehat{S}_{L}(d) = \left\{X_p; \left(\frac{1}{L}\sum_{l=1}^L 1\left(R_l(X_p)\le d\right)\right)>0\right\},
\end{equation}
which is exemplified in Panel (c) of Table \ref{tab:example}.

@@

\begin{theorem}
  Assume that $\{R_l(X)\}_{l=1}^L$ are independent draws from a
  probability distribution $Q$ on the set of lists.  Then,
  $\left\|\widehat{\textrm{sra}}_L - \textrm{sra}\right\|_\infty =
  o_P(1)$.
\label{thm:consistency}
\end{theorem}

We now define the parameter of interest. For any given deterministic threshold function $q\colon\, 1,\ldots,P \mapsto \mathbb{R}_{\geq 0}$
we define the population superlevel set of the sequential rank agreement as
\begin{align}
  \mathcal{L}(q) = \left\{d : \textrm{sra}(d) \geq q(d)\right\}.
\end{align}
The values of $q$ could for example be the limits-of-agreement for randomly
permuted lists as in the null-hypothesis in equation (\ref{eq:permutationHypothesis}).
A change-point $d^\ast(q)$ in the list agreement is then defined by the position
\begin{align}
  d^\ast(q) &= 1\left(|\mathcal{L}(q)| > 0\right)\inf \mathcal{L}(q) + 1\left(|\mathcal{L}(q)| = 0\right)P
\end{align}
corresponding to the first list depth where the sequential rank agreement exceeds the threshold 
if such a position exists. Otherwise, the full list is in agreement according to $q$ and the change-point
is defined as the length of the lists. The empirical superlevel set is similarly defined as
\begin{align}
 \widehat{\mathcal{L}}_L(\widehat{q}_L) = \left\{d : \widehat{\textrm{sra}}_L(d) \geq \widehat{q}_L(d)\right\}
\end{align}
where we allow the threshold function to depend on the sample size as well. The estimated change-point is therefore
\begin{align}
  \widehat{d^\ast_L}(\widehat{q}_L) &= 1(|\widehat{\mathcal{L}}_L(\widehat{q}_L)| > 0)\inf \widehat{\mathcal{L}}_L(\widehat{q}_L) + 1(|\widehat{\mathcal{L}}_L(\widehat{q}_L)| = 0)P.
\end{align}

The consistency of the estimated change-point, $\widehat{d^\ast_L}(\widehat{q}_L)$, follows from Theorem \ref{thm:consistency} by the following corollary. 
\begin{corollary}\label{thecor}
Let $\widehat{q}_L$ be a threshold function such that $\left\|\widehat{q}_L - q\right\|_\infty = o_P(1)$ for some limiting
function $q$.  Then, $\widehat{d^\ast_L}(\widehat{q}_L) \overset{P}{\longrightarrow} d^\ast(q)$ for $L \rightarrow \infty$.
\label{cor:consistency}
\end{corollary}

Corollary \ref{thecor} tells us that we can use the a threshold
function $\widehat{q}_L$ estimated under the null hypothesis as
discussed in the previous section as a limiting threshold function for
inferring the depth $d$, where the observed sequential rank agreement
first crosses the threshold of the null threshold, \ie, the depth
until which the observed ranked lists are in better agreement than
expected under the null hypothesis. In that sense the threshold
function serves the same role as the limits of agreement in method
comparison studies, except that the threshold funciton is not constant
but can accommodate the changing nature of the number of items used
for the computation of the equential rank agreement for a given depth.

@@

\mcomment{Elaborate more on how to apply this in practice (finite
  sample/bootstrap/etc.) and that $q$ is similar to LoA BA-style?  }

@@

\newpage
\section{Proof of Theorem \ref{thm:consistency}}

@@

\section{Proof of Corollary \ref{cor:consistency}}
Let $A \triangle B = (A \setminus B) \cup (B \setminus A)$ is the symmetric difference between sets $A$ and $B$. It is then sufficient to show that
$|\widehat{\mathcal{L}}_L(\widehat{q}_L) \triangle \mathcal{L}(q)|$ converges to zero in probability as $L \rightarrow \infty$. By the triangle inequality
we have that
\begin{align}
 \left|\widehat{\mathcal{L}}_L(\widehat{q}_L) \triangle \mathcal{L}(q)\right| &= \left|\left(\widehat{\mathcal{L}}_L(\widehat{q}_L) \triangle \widehat{\mathcal{L}}_L(q)\right) \cup \left(\widehat{\mathcal{L}}_L(q) \triangle \mathcal{L}(q)\right)\right|\label{eq:corSetDiff}\\
   &\leq \left|\widehat{\mathcal{L}}_L(q) \triangle \mathcal{L}(q)\right| + \left|\widehat{\mathcal{L}}_L(\widehat{q}_L) \triangle \widehat{\mathcal{L}}_L(q)\right|\nn
\end{align}
Looking at the first term we have
\begin{align}
  \left|\widehat{\mathcal{L}}_L(q) \triangle \mathcal{L}(q)\right| &= \sum_{d=1}^P \left|1\left(\widehat{\textrm{sra}}_L(d) \geq q(d)\right)  - 1\left(\textrm{sra}(d)  \geq q(d)\right)\right|\\
   &=
\end{align}
Similarly we have
\begin{align}
\left|\widehat{\mathcal{L}}_L(\widehat{q}_L) \triangle \widehat{\mathcal{L}}_L(q)\right| &= \sum_{d=1}^P \left|1\left(\widehat{\textrm{sra}}_L(d) \geq \widehat{q}_L(d)\right) - 1\left(\widehat{\textrm{sra}}_L(d) \geq q(d)\right)\right|\\
   &=
\end{align}

>>>>

The agreement of the lists regarding the rank given to an item $X_p$
can be measured by the variance across the lists
\begin{align}
  A(X_p) &= \E_Q\left[\left(R(X_p)-\E_Q R(X_p)\right)^2\right]\\
   &= \sum_{r\in\Pi} \left(r(X_p)-\E_Q R(X_p)\right)^2 Q(r),\nn
\end{align}
where $\Pi$ is the set of all permutations of $X$, $Q$ is a
probability mass function on $\Pi$, and
$\E_Q R(X_p)=\sum_{r\in\Pi} r(X_p)Q(r)$. The empirical counterpart is
\begin{align}
\widehat{A}_L(X_p) = \frac{1}{L-1}\sum_{i=1}^L (R_i(X_p) - \overline{R}_L(X_p))^2, \quad \overline{R}_L(X_p) = \frac{1}{L}\sum_{i=1}^L R_i(X_p).\label{eq:empDistance}
\end{align}
For each item, the function $\widehat{A}_L$ has an interpretation as the expected
Euclidean distance of the individual rankings from the expected
ranking over the $L$ lists, and it corresponds to the same measure
that is used in method comparison studies to compute the limits of
agreement \citep{alt:bland:1983}.

For an integer $1\le d\le P$ we define the expected set of unique
items found by merging the first $d$ elements across the possible
lists:
\begin{align}
S(d) &= \left\{X_p; \left(\sum_{r\in\Pi} 1\left(r(X_p)\le d\right) Q(r)\right) >0\right\}\label{eq:sumset}
\end{align}
where $1(\cdot)$ denotes the indicator function.
The empirical counterpart is the set of unique items ranked less than or
equal to $d$ in any of the $L$ lists:
\begin{equation}
\widehat{S}_{L}(d) = \left\{X_p; \left(\frac{1}{L}\sum_{l=1}^L 1\left(R_l(X_p)\le d\right)\right)>0\right\},
\end{equation}
which is exemplified in Panel (c) of Table \ref{tab:example}.

@@

\begin{theorem}
  Assume that $\{R_l(X)\}_{l=1}^L$ are independent draws from a
  probability distribution $Q$ on the set of lists $\Pi$. Then,
  $\left\|\widehat{\textrm{sra}}_L - \textrm{sra}\right\|_\infty =
  o_P(1)$.
\label{thm:consistency}
\end{theorem}
\proof See Appendix \ref{sec:appA}.

We now define the change point as the first crossing point of the
sequential rank agreement and a threshold function
$q\colon\,\{1,\ldots,P\} \mapsto \mathbb{R}_{\geq 0}$. The values of
$q$ could be a deterministic constant, or in the experimental setting,
for example, the limits-of-agreement obtained in randomly permuted
lists corresponding to the null-hypothesis in equation
(\ref{eq:permutationHypothesis}). We define the superlevel set of the
sequential rank agreement with respect to $q$ as
\begin{align}
  \mathcal{L}(q) = \left\{d : \textrm{sra}(d) \geq q(d)\right\}.
\end{align}
A change-point $d^\ast(q)$ in the list agreement is then defined by the position
\begin{equation}
  d^\ast(q) =
  \begin{cases}
    \inf(\mathcal{L}(q)) & |\mathcal{L}(q)| > 0\\
  P & |\mathcal{L}(q)| = 0
  \end{cases}
\end{equation}
corresponding to the first list depth where the sequential rank
agreement exceeds the threshold if such a position exists. Otherwise,
the full list is in agreement according to $q$ and the change-point is
set to the length of the lists. The empirical superlevel set is
similarly defined as
\begin{align}
 \widehat{\mathcal{L}}_L(\widehat{q}_L) = \left\{d : \widehat{\textrm{sra}}_L(d) \geq \widehat{q}_L(d)\right\}
\end{align}
where we allow the threshold function to depend on the sample size as
well. The estimated change-point is therefore
\begin{align}
  \widehat{d^\ast_L}(\widehat{q}_L) &= 1(|\widehat{\mathcal{L}}_L(\widehat{q}_L)| > 0)\inf \widehat{\mathcal{L}}_L(\widehat{q}_L) + 1(|\widehat{\mathcal{L}}_L(\widehat{q}_L)| = 0)P.
\end{align}

The consistency of the estimated change-point, $\widehat{d^\ast_L}(\widehat{q}_L)$, follows from Theorem \ref{thm:consistency} by the following corollary. 
\begin{corollary}\label{thecor}
Let $\widehat{q}_L$ be a threshold function such that $\left\|\widehat{q}_L - q\right\|_\infty = o_P(1)$ for some limiting
function $q$.  Then, $\widehat{d^\ast_L}(\widehat{q}_L) \overset{P}{\longrightarrow} d^\ast(q)$ for $L \rightarrow \infty$.
\label{cor:consistency}
\end{corollary}
\proof See Appendix \ref{sec:appB}.

Corollary \ref{thecor} indicates that we can use the a threshold
function $\widehat{q}_L$ estimated under the null hypothesis as
discussed in the previous section as a limiting threshold function for
inferring the depth $d$, where the observed sequential rank agreement
first crosses the threshold of the null threshold, \ie, the depth
until which the observed ranked lists are in better agreement than
expected under the null hypothesis. In that sense the threshold
function serves the same role as the limits of agreement in method
comparison studies, except that the threshold funciton is not constant
but can accommodate the changing nature of the number of items used
for the computation of the equential rank agreement for a given depth.

@@

% \mcomment{Elaborate more on how to apply this in practice (finite
  % sample/bootstrap/etc.) and that $q$ is similar to LoA BA-style?  }

@@

\newpage
\section{Proof of Theorem \ref{thm:consistency}}
\label{sec:appA}

@@

\section{Proof of Corollary \ref{cor:consistency}}
\label{sec:appB}

Recall that $\mathcal{L}$ is the superlevel set of list positions
where the sequential rank agreement exceeds the threshold function
$q$. Let $A \triangle B = (A \setminus B) \cup (B \setminus A)$ is the
symmetric difference between sets $A$ and $B$. It is then sufficient
to show that
$|\widehat{\mathcal{L}}_L(\widehat{q}_L) \triangle \mathcal{L}(q)|$
converges to zero in probability as $L \rightarrow \infty$. By the
triangle inequality we have that
\begin{align}
 \left|\widehat{\mathcal{L}}_L(\widehat{q}_L) \triangle \mathcal{L}(q)\right| &= \left|\left(\widehat{\mathcal{L}}_L(\widehat{q}_L) \triangle \widehat{\mathcal{L}}_L(q)\right) \cup \left(\widehat{\mathcal{L}}_L(q) \triangle \mathcal{L}(q)\right)\right|\label{eq:corSetDiff}\\
   &\leq \left|\widehat{\mathcal{L}}_L(q) \triangle \mathcal{L}(q)\right| + \left|\widehat{\mathcal{L}}_L(\widehat{q}_L) \triangle \widehat{\mathcal{L}}_L(q)\right|\nn
\end{align}
Looking at the first term we have
\begin{align}
  \left|\widehat{\mathcal{L}}_L(q) \triangle \mathcal{L}(q)\right| &= \sum_{d=1}^P \left|1\left(\widehat{\textrm{sra}}_L(d) \geq q(d)\right)  - 1\left(\textrm{sra}(d)  \geq q(d)\right)\right|\\
   &=
\end{align}
Similarly we have
\begin{align}
\left|\widehat{\mathcal{L}}_L(\widehat{q}_L) \triangle \widehat{\mathcal{L}}_L(q)\right| &= \sum_{d=1}^P \left|1\left(\widehat{\textrm{sra}}_L(d) \geq \widehat{q}_L(d)\right) - 1\left(\widehat{\textrm{sra}}_L(d) \geq q(d)\right)\right|\\
   &=
\end{align}
